\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-21 jasa/paper-jasa.tex
continued

% 2) Likelihood ratio tests
%
% Introduce the mathematical problem with notations (was 1.1 and 1.2 before)
% I would keep some of the comments of 1.3, but I feel some are not necessary

@@

% \subsection{Comments on classification and frequentist hypothesis tests}
%
% Vast literature exists around generative and discriminative
% classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
% for the joint probability $p(x,y)$, of the inputs $x$ and the classification
% label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative
% classifiers model the posterior $p(y|x)$ directly. For classification tasks, one
% then thresholds on $p(y|x)$. In both cases this description in terms of a
% posterior requires a prior distribution for $p(y)$, which is either modeled
% explicitly or learned from the training data. This familiar formulation of
% classification may lead to some confusion in the setting of the current work.
%
% The first possible source of confusion we wish to avoid is that here
% $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$,
% not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
% traditional scientific theory or simulator, able to make predictions about $x$
% and being motivated by domain-specific considerations.
%
% The second possible source of confusion is that we are not directly interested
% in calibrating the classification score in terms of a per-event posterior
% probability $p(y|x)$. Instead, we are interested in the approximation of the
% per-experiment likelihood function or likelihood ratio, which might be used for
% several purposes, including the calculation of p-values.
%
% Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
% we do not have a prior $\pi(\theta)$. While we can use the generative models to
% produce training data $(x,y=0)$ generated from $p(x|\theta_0)$ and $(x,y=1)$
% generated from $p(x|\theta_1)$, the relative mix $p(y)$ is arbitrary.  When
% $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
% \begin{equation}
% p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
% \end{equation}
% which is monotonic with the desired likelihood ratio
% $p(x|\theta_1)/p(x|\theta_0)$. Since the prior $p(y)$ is not needed for the
% target likelihood ratio test and because the classifier score $p(y|x)$ may not
% be well calibrated, we choose to denote the classifier score $s(x)$ and simply
% think of it as a deterministic dimensionality reduction map $s: X \to
% \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott} and
% \cite{Neal:2007zz}.

@@

The main result of this paper is to generalize the observation that one can
form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{Y} = s(\mathbf{X}; \theta_0, \theta_1)$ is based
on a parametrized function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x}; \theta_0, \theta_1)$ as a discriminative classifier.  In
Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier  parametrized in terms of $(\theta_0, \theta_1)$.

\subsection{Likelihood ratios under change of variable}

@@

@@

\glnote{Discuss that training on signal vs background instead of
signal+background vs background is OK because this learns a function which is
1:1.}

\glnote{Discuss the capacity of the classifier; Generalization to decomposed
ratios Discuss about the need of calibration in case of imperfect classifiers
(was first paragraphs of section 5)}

% \section{Embedding the classifier into the likelihood}
%
% Thus far we have shown that the target likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of $p(x|\theta_0)/p(x|\theta_1)$. We now generalize from the ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are continuous model parameters.
%  We postpone the practicalities of training the classifier and estimating the density to Section~\ref{S:classifier} and continue in the likelihood-free setting with idealized classifiers and their densities.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$ for the observed data and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0, \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the value of the parameters  that are being compared in the likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
% This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.

@@

@@

It is not obvious that if we are working with the distributions $p(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators.
Fortunately, there is a construction based on $p(s|\theta)$ that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
\begin{equation}\label{eq:mle_withs}
\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $p(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

\section{Learning the correct mapping and its distribution}\label{S:classifier}

In order for this approach to be useful in the likelihood-free setting, where it
is not possible to evaluate the density $p(x|\theta)$, we need to be able to
approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples
$\{x_i\}$ drawn from the generative model $p(x|\theta)$.  Denote the
dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its
distribution $\hat{p}(\hat{s}|\theta)$.

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x;
\theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta)
\approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor job of
reproducing the level sets of the per-event likelihood ratio, the density of
$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
the resulting inference will still be valid. This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

\subsection{The fixed discriminative classification setting} For fixed
$\theta_0$ and $\theta_1$ we can generate large samples from each model and
train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate
training data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data
$(x_i , y_i=1)$. With balanced training data
\mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the
desired per-event likelihood ratio $q(x)$. Thus, standard supervised learning
algorithms with various surrogate loss functions lead to discriminative
classifiers that approximate a monotonic function of per-event likelihood ratio
$q(x)$.

@@

% - typical usage in HEP with a concrete experimental example.
%   * illustrate the method for several classifiers (new)
%    * illustrate the method for several calibration routines (new)
%    * compare with other methods (new)
% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)

>>>>

@@

@@

The main result of this paper (see Theorem~\ref{thm:ratio-equivalence}) is to
generalize the observation that one can form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{Y} = s(\mathbf{X}; \theta_0, \theta_1)$ is based
on a parametrized function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x}; \theta_0, \theta_1)$ as a discriminative classifier.  In
Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier  parametrized in terms of $(\theta_0, \theta_1)$.

\subsection{Likelihood ratios under change of variables}

@@

% \subsection{The fixed discriminative classification setting} For fixed
% $\theta_0$ and $\theta_1$ we can generate large samples from each model and
% train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate
% training data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data
% $(x_i , y_i=1)$. With balanced training data
% \mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
% classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
% p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the
% desired per-event likelihood ratio $q(x)$. Thus, standard supervised learning
% algorithms with various surrogate loss functions lead to discriminative
% classifiers that approximate a monotonic function of per-event likelihood ratio
% $q(x)$.

@@

\subsection{Learning the correct mapping and its distribution}

\glnote{This section needs to be improved.}

In order for this approach to be useful in the likelihood-free setting, where it
is not possible to evaluate the density $p(\mathbf{x}|\theta)$, we need to be
able to approximate both $s(\mathbf{x};\theta_0, \theta_1)$ and
$p(s(\mathbf{x})|\theta)$ based on samples $\{x_i\}$ drawn from the generative
model $p(\mathbf{x}|\theta)$.  Denote the dimensionality reduction map
$\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}; \theta_0, \theta_1) \approx
s(\mathbf{x}; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat
s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor
job of reproducing the level sets of the per-event likelihood ratio, the density
of $\hat{s}$ can still be well calibrated. In that case, one might loose power,
but the resulting inference will still be valid. This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

\subsection{Classification and frequentist hypothesis tests}

\glnote{Shall we keep everything from this?}

Vast literature exists around generative and discriminative
classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
for the joint probability $p(\mathbf{x}, y)$, of the inputs $x$ and the classification
label $y$, and predict $p(y|\mathbf{x})$ via Bayes rule. In contrast, discriminative
classifiers model the posterior $p(y|\mathbf{x})$ directly. For classification tasks, one
then thresholds on $p(y|\mathbf{x})$. In both cases this description in terms of a
posterior requires a prior distribution for $p(y)$, which is either modeled
explicitly or learned from the training data. This familiar formulation of
classification may lead to some confusion in the setting of the current work.

The first possible source of confusion we wish to avoid is that here
$p(\mathbf{x}|\theta)$  is a  \textit{generative statistical model} for the features $\mathbf{x}$,
not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
traditional scientific theory or simulator, able to make predictions about $\mathbf{x}$
and being motivated by domain-specific considerations.

The second possible source of confusion is that we are not directly interested
in calibrating the classification score in terms of a per-event posterior
probability $p(y|\mathbf{x})$. Instead, we are interested in the approximation of the
per-experiment likelihood function or likelihood ratio, which might be used for
several purposes, including the calculation of p-values.

Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
we do not have a prior $\pi(\theta)$. While we can use the generative models to
produce training data $(\mathbf{x} ,y=0)$ generated from
$p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, y=1)$ generated from
$p(\mathbf{x}|\theta_1)$, the relative mix $p(y)$ is arbitrary. Since the prior
$p(y)$ is not needed for the target likelihood ratio test and because the
classifier score $p(y|\mathbf{x})$ may not be well calibrated, we choose to denote the
classifier score $s(\mathbf{x})$ and simply think of it as a deterministic dimensionality
reduction map $s: \mathbb{R}^p \to \mathbb{R}$.  Similar points have been made
by~\cite{ClaytonScott} and \cite{Neal:2007zz}.

@@

% Thus far we have shown that the target likelihood ratio
% $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be
% reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the
% classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of
% $p(x|\theta_0)/p(x|\theta_1)$. We now generalize from the ratio of two simple
% hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are
% continuous model parameters. We postpone the practicalities of training the
% classifier and estimating the density to Section~\ref{S:classifier} and continue
% in the likelihood-free setting with idealized classifiers and their densities.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
% for the observed data and never refer back to the original features $x_e$. In
% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the
% classifier to the stage of evaluating the enveloping likelihood ratio, then we
% can identify the value of the parameters  that are being compared in the
% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
%
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
%
% This is equivalent to approximating the likelihood function for $\theta_0$  when
% $\theta_1$ is held fixed.

@@

It is not obvious that if we are working with the distributions $p(s|\theta)$
(for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the
same estimators. Fortunately, there is a construction based on $p(s|\theta)$
that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$
for some fixed value of $\theta_1$. This allows us to use Theorem~1 toreformulate the maximum likelihood estimate

\begin{equation}\label{eq:mle_withs}
\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $p(s(x_e; \theta, \theta_1)
|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

@@

% - typical usage in HEP with a concrete experimental example.
%   * illustrate the method for several classifiers (new)
%    * illustrate the method for several calibration routines (new)
%    * compare with other methods (new)
% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
%   discuss about focusing the capacity of the classifier
% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)
